<!-- Shared body of the genetic gravity walkthrough.  Included by
     vignettes/genetic_gravity.Rmd (knitr child) and docs/genetic-gravity.qmd
     (Quarto include); edit here so the vignette and the website stay identical. -->

A Population Graph is undirected: each edge is a single, symmetric conditional
genetic distance between two populations.  *Genetic gravity* reads direction
out of that symmetric graph by asking, for every edge, how near it looks from
each of its two ends.  The few quantities involved are set out first, followed by a complete analysis, from raw genotypes to a Population Graph,
source-sink scores, tests for a directional gradient and a map of the gravity
field.

```{r libraries, message = FALSE}
library(gstudio)
library(dplyr)
library(ggplot2)
```

The code below is written as pipelines, with each function on its own line.
Every gstudio function takes its data as the first argument, so the same calls
also work in the older style, with options such as `stratum =` passed as
arguments; those versions are collected at the end, under *Without pipes*.

## The model in brief

Let $e_{ij}$ be the weight (conditional genetic distance, cGD) of the edge
between populations $i$ and $j$, $N(i)$ the populations connected to $i$,
$k_i = |N(i)|$ its degree and $s_i = \sum_{j \in N(i)} e_{ij}$ its strength.

**Neighbourhood weights.**  Population $j$'s weight in $i$'s neighbourhood is a
Gaussian kernel over $i$'s retained neighbours only, so by the Markov
properties of the graph nothing beyond them contributes:

$$
w_{j\mid i} = \frac{\exp\left(-e_{ij}^2 / 2b_i^2\right)}
                   {\sum_{k \in N(i)} \exp\left(-e_{ik}^2 / 2b_i^2\right)},
\qquad b_i = \gamma \, \frac{s_i}{k_i}.
$$

The bandwidth $b_i$ is a multiple $\gamma$ of the mean edge weight around $i$,
so rescaling every edge weight leaves everything below unchanged.

**Gravity asymmetry.**  Every edge is read from both ends.  The signed difference

$$
\Delta_{i \to j} = w_{i\mid j} - w_{j\mid i}
$$

is $i$'s pull on $j$ minus $j$'s pull on $i$.  Positive values make $i$ the
genetic source and $j$ the sink; throughout, **higher means more of a source**.

**Gravity and source-sink scores.**  The gravity of population $i$ is the total
weight it holds in its neighbours' neighbourhoods, $g_i = \sum_{j \in N(i)} w_{i\mid j}$.
Because each neighbourhood sums to one, $\sum_i g_i = K$, so the balance
$g_i - 1$ sums to zero: every source is offset by sinks.  The source-sink score
$S$ puts all populations on one scale, as the values whose differences best
reproduce the asymmetries on every edge:

$$
\hat{S} = \arg\min_S \sum_{(i,j) \in E} \left[\Delta_{i \to j} - (S_i - S_j)\right]^2,
\qquad \textstyle\sum_i S_i = 0 \text{ within each connected component.}
$$

**Degree and the bandwidth.**  As the kernel flattens, $w_{j\mid i} \to 1/k_i$
and the asymmetry falls to a floor set by degree alone,
$\Delta^0_{i \to j} = 1/k_j - 1/k_i$.  The share of the variance in $\Delta$
explained by $\Delta^0$ (the *degree share*) reports how much of an analysis is
topology rather than genetics.  Two bandwidths answer different questions:
$\gamma = 1/2$ (the default) is *degree-neutral* and is used for inferring
direction; $\gamma = 1$, the local mean, keeps the contribution of degree and
describes each population's overall pull in the network.

**Partitioned distance.**  The same weights split each edge into two directed
shares, the partitioned conditional genetic distance (pGD):

$$
p_{ij} = e_{ij} \, \frac{w_{j\mid i}}{w_{j\mid i} + w_{i\mid j}}, \qquad p_{ij} + p_{ji} = e_{ij}.
$$

Being a distance, pGD is *shorter* in the direction of gene flow.  It turns the
undirected graph into a directed one for downstream analyses such as isolation
by graph distance.

## The data

`gravity` holds genotypes from one census of an individual-based simulation:
25 populations of 100 diploids on a one-dimensional stepping stone, scored at
20 biallelic loci.  Gene flow runs four times more strongly toward
higher-numbered populations than back (`Pop01` is the upstream end), so the
answer is known in advance.  See `?gravity` for how the census was chosen.

```{r data}
data(gravity)
dim(gravity)

gravity |>
  select(Population, L01:L04) |>
  head(4)
```

## From genotypes to a Population Graph

`population_graph()` encodes the genotypes and builds the graph among the
strata in the `Population` column (its default `stratum`).

```{r graph}
graph <- gravity |>
  population_graph()

graph
```

The hypothesized axis of flow is position along the stepping stone, which here
is the population number:

```{r deme}
nodes <- igraph::V(graph)$name

deme <- nodes |>
  sub(pattern = "Pop", replacement = "") |>
  as.integer() |>
  setNames(nodes)
```

`plot()` draws the graph.  Without coordinates it uses a Kamada-Kawai layout.
`population_graph()` stores each population's within-population genetic
variability as the `size` vertex attribute, and nodes are sized by it.

```{r plot-graph}
graph |>
  plot()
```

## Neighbourhood weights and edge asymmetries

`neighbourhood_weights()` returns $W_{ij} = w_{j\mid i}$; each row sums to one
over that population's neighbours.

```{r weights}
W <- graph |>
  neighbourhood_weights()

round(W[1:5, 1:5], 3)
rowSums(W, na.rm = TRUE)[1:5]
```

`gravity_edges()` gives one row per edge with both weights, $\Delta$ and the
degree-only floor $\Delta^0$.  Supplying the covariate `x` orients every edge
from the lower to the higher value of `x`, so a positive `Delta` means the
upstream population is the source.

```{r edges}
edges <- graph |>
  gravity_edges(x = deme)

edges |>
  head()

edges |>
  summarize(edges = n(),
            upstream_source = mean(Delta > 0))
```

## Source-sink scores and the bandwidth

`source_sink_scores()` returns each population's gravity, balance and $S$, with
the degree diagnostics attached.

```{r scores}
scores <- graph |>
  source_sink_scores(x = deme)

scores |>
  select(Stratum, S, gravity, balance, degree) |>
  head()

attr(scores, "degree_share")
```

Plotted against position, $S$ falls steadily downstream: sources upstream,
sinks downstream.

```{r scores-plot, fig.height = 3.5}
scores |>
  mutate(deme = deme[Stratum]) |>
  ggplot(aes(deme, S)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point(aes(size = degree)) +
  labs(x = "Position along the stepping stone (upstream to downstream)",
       y = "Source-sink score S") +
  theme_minimal()
```

The degree share says how much of $\Delta$ is set by the number of neighbours.
At the degree-neutral bandwidth it is small; at the local mean, degree carries
more of the signal.  A gravity field computed at both bandwidths reports the
diagnostics side by side:

```{r bandwidth}
graph |>
  gravity_field(gamma = c(0.5, 1)) |>
  as.data.frame(what = "diagnostics") |>
  select(gamma, degree_share, cor_S_degree, gradient_share)
```

Report the bandwidth and the degree share with any directional result.

## Testing for a directional gradient

Two complementary procedures test for direction.  Neither permutes parts of
the graph freely: $\Delta$ and $S$ inherit smooth, node-level structure from
drift on the graph, and nulls that ignore it are anti-conservative.

### Source-sink gradient test

The statistic is Spearman's $r(S, x)$.  The null is Moran spectral
randomization of $S$ on the graph itself, which reproduces the smoothness $S$
inherits from the topology.  A negative $r$ places the sources at low $x$.

```{r ss-test}
set.seed(1)

graph |>
  source_sink_test(x = deme)
```

### Directional isolation by graph distance

The second procedure asks whether modelling direction improves isolation by
graph distance.  It regresses all-pairs cGD on spatial separation ($R^2_{cGD}$),
then regresses all-pairs directed pGD on separation with separate forward and
reverse slopes ($R^2_{pGD}$, adjusted).  $\Delta R^2 = R^2_{pGD} - R^2_{cGD} > 0$
means the directional description fits better; the forward-to-reverse slope
ratio gives the direction and strength (below one: genetic distance
accumulates more slowly downstream).  This is not a nested-model test, and
$\Delta R^2 \le 0$ is an absence of evidence for asymmetry, not evidence of
symmetry.

`ibgd()` does both steps and prints each as a standard test.  The default,
`mode = "cgd"`, is ordinary isolation by graph distance, with a Mantel
permutation test of the correlation between cGD and separation;
`mode = "pgd"` adds the directional model and reports $\Delta R^2$ with the
forward and reverse slopes.  It gives no p-value for $\Delta R^2$, because the
two fits are not nested.

```{r ibgd}
graph |>
  ibgd(x = deme)

graph |>
  ibgd(x = deme, mode = "pgd")
```

Both procedures recover the imposed direction: a strong negative gradient in
$S$ (sources upstream), and a directional isolation model that fits better than
the symmetric one with far shallower downstream than upstream isolation.

pGD is available directly for other downstream analyses.  From the genotypes,
`genetic_distance()` returns all-pairs directed shortest-path pGD alongside the
other distance metrics, where `P[i, j]` is the distance from $i$ to $j$; from an
existing graph, `pgd()` gives the same matrix (`output = "matrix"`) or the
directed graph of edge shares (`output = "graph"`).

```{r pgd}
P <- gravity |>
  genetic_distance(mode = "pgd")

round(P[1:4, 1:4], 3)
```

## Mapping the gravity field

`gravity_field()` collects every population-level and edge-level quantity;
`plot()` interpolates $S$ over the node positions (red = source, blue = sink)
and draws an arrow on each edge from source to sink, with length proportional
to $|\Delta|$.  Node placement is a drawing choice made in `plot()`; without
coordinates, layouts are computed on the directed pGD graph so that gene flow
shapes the picture.

```{r field}
field <- graph |>
  gravity_field()

field

field |>
  plot()
```

Fields with several panels are drawn on common scales.  Here the same graph at
the local-mean and degree-neutral bandwidths:

```{r field-gamma, fig.width = 9, fig.height = 5}
graph |>
  gravity_field(gamma = c(1, 0.5)) |>
  plot(node_labels = "none")
```

All the quantities behind the map are available as tidy tables for your own
figures:

```{r field-tables}
field |>
  as.data.frame(what = "edges") |>
  arrange(desc(abs(delta))) |>
  head()
```

## Default behaviour on a real graph

Nothing above is specific to simulated data.  For an existing Population Graph
the defaults give a complete picture in one line, here for `lopho`, the
Population Graph of the Sonoran Desert cactus *Lophocereus*.

```{r lopho}
data(lopho)

lopho |>
  gravity_field() |>
  plot()
```

## Without pipes

The pipelines above are the recommended style.  Every call also works with its
options passed as arguments, without dplyr; these are the equivalents of the
steps above, shown here but not run:

```{r no-pipes}
#| eval: false
graph  <- popgraph(to_mv(gravity), groups = gravity$Population)
deme   <- as.integer(sub("Pop", "", igraph::V(graph)$name))
names(deme) <- igraph::V(graph)$name

W      <- neighbourhood_weights(graph, gamma = 0.5)
edges  <- gravity_edges(graph, gamma = 0.5, x = deme)
scores <- source_sink_scores(graph, gamma = 0.5, x = deme)

source_sink_test(graph, x = deme, gamma = 0.5, nperm = 999)
ibgd(graph, x = deme, mode = "cgd", nperm = 999)
ibgd(graph, x = deme, mode = "pgd", gamma = 0.5)

P      <- genetic_distance(gravity, stratum = "Population", mode = "pgd", gamma = 0.5)
field  <- gravity_field(graph, gamma = 0.5)
plot(field, layout = "kk", surface = "interpolate")
```
