# Allele Frequencies & HWE

This chapter covers estimating allele and genotype frequencies, testing for Hardy-Weinberg Equilibrium, and rarefaction analysis.

```{r}
#| message: false
library(gstudio)
library(dplyr)
library(ggplot2)
data(arapat)
```

## Allele Frequencies

The `frequencies()` function is an S3 generic that works on both individual locus vectors and entire data.frames.

### Single Locus

```{r}
arapat |>
  pull(LTRS) |>
  frequencies()
```

### Entire Data Frame

When applied to a data.frame, `frequencies()` returns frequencies for all locus columns:

```{r}
freqs <- arapat |>
  frequencies()

freqs |>
  head(12)
```

### By Stratum

Add a `stratum` argument to partition frequencies by population:

```{r}
freqs_pop <- arapat |>
  frequencies(stratum = "Species")

freqs_pop |>
  head(12)
```

Because the result is a tidy data frame, dplyr summaries follow directly, for
example the most common allele at each locus in each species:

```{r}
freqs_pop |>
  group_by(Stratum, Locus) |>
  slice_max(Frequency, n = 1, with_ties = FALSE) |>
  filter(Locus %in% c("LTRS", "WNT"))
```

### Selecting Specific Loci

```{r}
arapat |>
  frequencies(loci = c("LTRS", "WNT"))
```

## Frequency Matrix

The `frequency_matrix()` function produces a wide-format matrix of allele frequencies:

```{r}
fm <- arapat |>
  frequency_matrix(stratum = "Species")

fm[, 1:8]
```

## Genotype Counts and Frequencies

### Genotype Counts per Stratum

The `genotype_counts()` function summarizes non-missing genotype counts per locus, optionally by stratum:

```{r}
arapat |>
  genotype_counts()
```

```{r}
arapat |>
  genotype_counts(stratum = "Species")
```

### Genotype Frequencies

The `genotype_frequencies()` function returns observed and expected genotype counts for a locus vector:

```{r}
arapat |>
  pull(LTRS) |>
  genotype_frequencies()
```

## Hardy-Weinberg Equilibrium

The `hwe()` function tests for HWE using a chi-square approximation:

```{r}
arapat |>
  filter(Species == "Mainland") |>
  hwe()
```

The output includes the chi-square statistic, degrees of freedom, and p-value for each locus. Significant p-values indicate departures from HWE expectations.

### Interpreting Results

Departures from HWE can arise from:

- Non-random mating (inbreeding, assortative mating)
- Population substructure (Wahlund effect)
- Selection at or near the marker locus
- Small population size (genetic drift)

## Rarefaction

The `rarefaction()` function subsamples the data repeatedly at a smaller sample size to estimate the distribution of a diversity statistic:

```{r}
rare_vals <- arapat |>
  pull(LTRS) |>
  rarefaction(mode = "Ae", size = 20, nperm = 199)

data.frame(Ae = rare_vals) |>
  ggplot(aes(Ae)) +
  geom_histogram(bins = 20, fill = "steelblue") +
  labs(title = "Rarefied Ae at LTRS", x = "Effective alleles", y = "Count") +
  theme_minimal()
```

This is useful for comparing diversity across populations with unequal sample sizes.

## Optimal Sampling

The `optimal_sampling()` function estimates how many individuals are needed to capture a given proportion of allelic diversity:

```{r}
#| eval: false
arapat |>
  pull(LTRS) |>
  optimal_sampling(nrep = 99)
```

## Without Pipes

```{r}
#| eval: false
frequencies(arapat$LTRS)
frequencies(arapat, stratum = "Species")
frequencies(arapat, loci = c("LTRS", "WNT"))
frequency_matrix(arapat, stratum = "Species")
genotype_counts(arapat, stratum = "Species")
genotype_frequencies(arapat$LTRS)
hwe(arapat[arapat$Species == "Mainland", ])
rare_vals <- rarefaction(arapat$LTRS, mode = "Ae", size = 20, nperm = 199)
hist(rare_vals)
optimal_sampling(arapat$LTRS, nrep = 99)
```
