# Importing & Exporting Data

This chapter covers reading genetic data from files, converting between formats, and generating synthetic populations.

```{r}
#| message: false
library(gstudio)
library(dplyr)
```

## Reading Population Data

The `read_population()` function is the primary data import tool. It reads text files and converts specified columns to `locus` objects.

### Function Signature

```{r}
#| eval: false
read_population(path, type, locus.columns, phased = FALSE,
                sep = ",", header = TRUE, delim = ":")
```

The key parameters:

- `path` --- file path to the data
- `type` --- marker encoding format
- `locus.columns` --- integer vector indicating which columns contain genetic data

### Separated Format

The most common format uses colon-separated alleles (e.g., `"A:B"`):

```{r}
data_sep <- system.file("extdata", "data_separated.csv", package = "gstudio") |>
  read_population(type = "separated", locus.columns = c(4, 5))

data_sep |>
  head()
```

### Column Format

Two adjacent columns per locus (one column per allele):

```{r}
data_col <- system.file("extdata", "data_2_column.csv", package = "gstudio") |>
  read_population(type = "column", locus.columns = c(4:7))

data_col |>
  head()
```

### SNP Format

Encoded as 0, 1, 2 (minor allele count):

```{r}
data_snp <- system.file("extdata", "data_snp.csv", package = "gstudio") |>
  read_population(type = "snp", locus.columns = c(4:7))

data_snp |>
  head()
```

### AFLP Format

Binary (0/1) presence-absence:

```{r}
data_aflp <- system.file("extdata", "data_aflp.csv", package = "gstudio") |>
  read_population(type = "aflp", locus.columns = c(4:7))

data_aflp |>
  head()
```

### Allozyme (Zyme) Format

Concatenated integer alleles (e.g., "12" for alleles 1 and 2):

```{r}
data_zyme <- system.file("extdata", "data_zymelike.csv", package = "gstudio") |>
  read_population(type = "zyme", locus.columns = c(4:7))

data_zyme |>
  head()
```

### Genepop Format

Genepop files have their own structure with `Pop` delimiters:

```{r}
data_gp <- system.file("extdata", "genepop.txt", package = "gstudio") |>
  read_population(type = "genepop")

data_gp |>
  head()
```

### Structure Format

Two-row-per-individual STRUCTURE format:

```{r}
data_str <- system.file("extdata", "data_structure.str", package = "gstudio") |>
  read_population(type = "structure")

data_str |>
  head()
```

## Writing Population Data

The `write_population()` function exports a data.frame to various formats:

```{r}
#| eval: false
# Plain text (colon-separated loci)
arapat |>
  write_population(file = "arapat.csv")

# Genepop format, for one species only
arapat |>
  filter(Species == "Cape") |>
  write_population(file = "cape.gen", mode = "genepop", stratum = "Population")

# Structure format
arapat |>
  write_population(file = "arapat.str", mode = "structure", stratum = "Species")
```

## Direct Format Conversion

You can also convert data.frames directly without writing to disk:

```{r}
data(arapat)

gp <- arapat |>
  to_genepop(stratum = "Species")

cat(gp[1:10], sep = "\n")
```

```{r}
st <- arapat |>
  to_structure(stratum = "Species")

cat(st[1:5], sep = "\n")
```

## Generating Synthetic Data

### From Allele Frequencies

The `make_population()` function creates a random population from a frequency table:

```{r}
freqs <- data.frame(
  Locus = rep(c("TPI", "PGM"), each = 3),
  Allele = c("A", "B", "C", "H", "I", "J"),
  Frequency = c(0.55, 0.30, 0.15, 0.34, 0.34, 0.32)
)
pop <- freqs |>
  make_population(N = 10)

pop |>
  head()
```

### Multiple Populations

The `make_populations()` function creates several populations at once from a frequency data.frame with a `Population` column:

```{r}
freqs <- expand.grid(Allele = c("01", "02"),
                     Locus = c("Loc1", "Loc2"),
                     Population = c("A", "B"),
                     stringsAsFactors = FALSE) |>
  mutate(Frequency = c(0.2, 0.8, 0.3, 0.7,
                       0.7, 0.3, 0.5, 0.5),
         Population = factor(Population, ordered = TRUE)) |>
  select(Population, Locus, Allele, Frequency)

pops <- freqs |>
  make_populations(N = 15)

pops |>
  head()

pops |>
  count(Population)
```

### From Individual Loci

The `make_loci()` function generates a vector of random locus objects:

```{r}
f <- data.frame(Allele = c("A", "B"), Frequency = c(0.6, 0.4))
loci <- f |>
  make_loci(N = 10)

loci
```

## Without Pipes

```{r}
#| eval: false
path     <- system.file("extdata", "data_separated.csv", package = "gstudio")
data_sep <- read_population(path, type = "separated", locus.columns = c(4, 5))
write_population(arapat, file = "arapat.str", mode = "structure", stratum = "Species")
gp   <- to_genepop(arapat, stratum = "Species")
pop  <- make_population(freqs, N = 10)
pops <- make_populations(freqs, N = 15)
loci <- make_loci(f, N = 10)
```
