# Genetic Diversity

This chapter covers estimation of genetic diversity statistics using the `genetic_diversity()` wrapper and its underlying per-locus functions.

```{r}
#| message: false
library(gstudio)
library(dplyr)
library(ggplot2)
data(arapat)
```

## The `genetic_diversity()` Wrapper

The `genetic_diversity()` function is the main entry point for diversity estimation. It dispatches to specific estimators based on the `mode` parameter:

| Mode | Statistic | Description |
|------|-----------|-------------|
| `"A"` | Allelic richness | Number of alleles |
| `"Ae"` | Effective alleles | $1 / \sum p_i^2$ |
| `"A95"` | A (5% threshold) | Alleles with frequency >= 0.05 |
| `"He"` | Expected heterozygosity | $1 - \sum p_i^2$ |
| `"Ho"` | Observed heterozygosity | Proportion of heterozygotes |
| `"Hes"` | Corrected He | Nei's small-sample corrected He |
| `"Hos"` | Corrected Ho | Nei's small-sample corrected Ho |
| `"Fis"` | Inbreeding coefficient | $1 - H_o/H_e$ |
| `"Pe"` | Polymorphic index | Whether the locus is polymorphic |

## Global (Unstratified) Estimates

Without a stratum, diversity is estimated across all individuals:

```{r}
arapat |>
  genetic_diversity(mode = "A")

arapat |>
  genetic_diversity(mode = "He")
```

Several statistics at once, joined into one table:

```{r}
c("A", "Ae", "He", "Ho") |>
  lapply(\(m) genetic_diversity(arapat, mode = m)) |>
  Reduce(f = \(x, y) inner_join(x, y, by = "Locus"))
```

## Per-Stratum Estimates

Provide a `stratum` argument to compute diversity within each population:

```{r}
arapat |>
  genetic_diversity(stratum = "Species", mode = "He")
```

```{r}
arapat |>
  genetic_diversity(stratum = "Species", mode = "Ae")
```

## Individual Estimator Functions

Each diversity metric has a standalone function that operates on a single locus vector.

### Allelic Richness

```{r}
arapat |>
  pull(LTRS) |>
  A()
```

### Effective Number of Alleles

```{r}
arapat |>
  pull(LTRS) |>
  Ae()
```

### Expected Heterozygosity

```{r}
arapat |>
  pull(LTRS) |>
  He()
```

### Observed Heterozygosity

```{r}
arapat |>
  pull(LTRS) |>
  Ho()
```

### Inbreeding Coefficient

```{r}
arapat |>
  pull(LTRS) |>
  Fis()
```

### Exclusion Probability

```{r}
arapat |>
  pull(LTRS) |>
  Pe()
```

## Corrected Heterozygosity (Nei's)

The `Hes` and `Hos` modes require a stratum because they apply a small-sample correction per subpopulation:

```{r}
arapat |>
  genetic_diversity(stratum = "Species", mode = "Hes")
```

## Multilocus Diversity

The `multilocus_diversity()` function computes a summary across all loci:

```{r}
arapat |>
  multilocus_diversity()
```

## Selecting Specific Loci

You can restrict the analysis to a subset of loci:

```{r}
arapat |>
  genetic_diversity(loci = c("LTRS", "WNT"), mode = "He")
```

or select the columns first:

```{r}
arapat |>
  select(Species, LTRS, WNT) |>
  genetic_diversity(stratum = "Species", mode = "He")
```

## Comparing Populations

A common workflow is to estimate diversity per population, then compare:

```{r}
he_by_pop <- arapat |>
  genetic_diversity(stratum = "Species", mode = "He")

he_by_pop |>
  group_by(Stratum) |>
  summarize(mean_He = mean(He), sd_He = sd(He))
```

```{r}
he_by_pop |>
  ggplot(aes(x = Stratum, y = He, fill = Stratum)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Expected Heterozygosity by Species",
       y = "He") +
  theme(legend.position = "none")
```

## Without Pipes

```{r}
#| eval: false
genetic_diversity(arapat, mode = "He")
genetic_diversity(arapat, stratum = "Species", mode = "He")
genetic_diversity(arapat, loci = c("LTRS", "WNT"), mode = "He")
He(arapat$LTRS)
multilocus_diversity(arapat)
he_by_pop <- genetic_diversity(arapat, stratum = "Species", mode = "He")
ggplot(he_by_pop, aes(x = Stratum, y = He, fill = Stratum)) +
  geom_boxplot()
```
