# Genetic Gravity {#sec-genetic-gravity}

<!-- The walkthrough below is shared with the package vignette
     (vignette("genetic_gravity")); edit it in vignettes/children/genetic_gravity_body.Rmd
     and vignettes/children/genetic_gravity_summary.Rmd, not here. -->

{{< include ../vignettes/children/genetic_gravity_body.Rmd >}}

## Empirical data: *Araptus attenuatus*

The same analysis applies directly to field data.  Here the Population Graph
is built from the `arapat` genotypes with `population_graph()`, and the
populations' geographic centroids give both the map layout and a covariate to
test against: latitude along the Baja California peninsula.

```{r arapat-graph}
data(arapat)

beetles <- arapat |>
  population_graph()

coords <- arapat |>
  strata_coordinates()

lat <- coords |>
  pull(Latitude, name = Stratum)
```

Passing the centroids as `layout` draws the gravity field on geographic
coordinates.  Sampling locations are clustered, so the surface is extended with
`mask_dist` (by default it reaches 1.2 times the median nearest-neighbour
distance between populations).  Comparing the two bandwidths shows how much of
the local-mean picture is degree:

```{r arapat-field}
#| fig-width: 9
#| fig-height: 5
both <- beetles |>
  gravity_field(gamma = c(1, 0.5))

both |>
  as.data.frame(what = "diagnostics")

both |>
  plot(layout = coords, mask_dist = 0.6)
```

Both tests take latitude as the hypothesized axis:

```{r arapat-tests}
set.seed(1)

beetles |>
  source_sink_test(x = lat, nperm = 499)

beetles |>
  ibgd(x = lat)

beetles |>
  ibgd(x = lat, mode = "pgd")
```

Without pipes, the same steps are:

```{r arapat-no-pipes}
#| eval: false
beetles   <- population_graph(arapat, stratum = "Population")
coords <- strata_coordinates(arapat, stratum = "Population")
lat    <- setNames(coords$Latitude, coords$Stratum)
plot(gravity_field(beetles, gamma = c(1, 0.5)), layout = coords, mask_dist = 0.6)
source_sink_test(beetles, x = lat, nperm = 499)
ibgd(beetles, x = lat, mode = "cgd")
ibgd(beetles, x = lat, mode = "pgd")
```

The two tests answer different questions.  Under strong drift the model
comparison can prefer a directional fit by chance, so read it alongside the
source-sink test.  Edge-level permutation tests of $\Delta$
(`asymmetry_significance()` and its modes) are not recommended as tests of
direction: under symmetric migration they are anti-conservative, or never
reject.

## Data quality: information loss

As loci fix, graphs are estimated from fewer informative loci, which inflates
$|\Delta|$ whatever the direction of gene flow.  `locus_subsample_null()`
rebuilds the graph from random subsets of polymorphic loci, to show how much
$|\Delta|$ information loss alone can produce at a given level of polymorphism:

```{r info-loss}
#| eval: false
arapat |>
  locus_subsample_null(k = c(2, 3, 4, 6), nsub = 3)
```

## Simulating directional gene flow

A directional stepping stone, with rates to the next population downstream
(`m_fwd`) and upstream (`m_rev`), plugs into `simulate_pop()`; this is how the
`gravity` data were generated (at a much larger scale):

```{r simulate}
#| eval: false
demes <- sprintf("Pop%02d", 1:10)

mm <- demes |>
  migration_matrix(model = "stepping_stone_1d_asym", m_fwd = 0.04, m_rev = 0.01)

out <- data.frame(Population = rep(demes, each = 2),
                  Locus = "L1",
                  Allele = rep(c("01", "02"), 10),
                  Frequency = 0.5) |>
  make_populations(N = 50) |>
  simulate_pop(ngen = 200, migration = migration_event(mm), verbose = FALSE)
```

{{< include ../vignettes/children/genetic_gravity_summary.Rmd >}}
