# Population Structure

This chapter covers the estimation of genetic structure statistics --- measures of how genetic variation is partitioned among populations.

```{r}
#| message: false
library(gstudio)
library(dplyr)
data(arapat)
```

## The `genetic_structure()` Wrapper

The `genetic_structure()` function dispatches to specific structure estimators based on the `mode` parameter:

| Mode | Statistic | Description |
|------|-----------|-------------|
| `"Gst"` | Nei's Gst | Proportion of total diversity among populations |
| `"Gst_prime"` | Hedrick's G'st | Gst corrected for marker diversity |
| `"Dest"` | Jost's D | True differentiation measure |
| `"Fst"` | Wright's Fst | Fixation index |

## Basic Usage

Estimate Gst across species:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst")
```

The output provides per-locus estimates plus a multilocus estimate (last row). The multilocus estimate follows the approach of Culley et al. (2001).

## Hedrick's Correction

For highly polymorphic markers, `Gst` is constrained to small values. Hedrick's correction accounts for this:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst_prime")
```

## Jost's D

An alternative differentiation measure that is independent of within-population diversity:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Dest")
```

## Permutation Testing

Use the `nperm` parameter to test the null hypothesis that the structure parameter equals zero:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst", nperm = 99)
```

The resulting P column indicates the probability of observing a value as extreme as the estimate under random permutation of individuals among strata.

## Pairwise Structure

Set `pairwise = TRUE` to compute structure between all pairs of strata:

```{r}
pw <- arapat |>
  genetic_structure(stratum = "Species", mode = "Gst", pairwise = TRUE)
pw
```

This returns a symmetric matrix of pairwise multilocus estimates.

## Restricting to Specific Loci

Use the `locus` parameter to restrict the analysis:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst", locus = "LTRS")
```

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst", locus = c("LTRS", "WNT"))
```

## Multilocus Assignment

The `multilocus_assignment()` function estimates the probability that an individual originates from each population based on multilocus genotype frequencies:

```{r}
# Compute stratum-level frequencies
freqs <- arapat |>
  frequencies(stratum = "Species")

# Assign the first individual
arapat |>
  slice(1) |>
  multilocus_assignment(freqs)
```

The `Posterior` column gives the relative probability of assignment to each stratum given the individual's multilocus genotype.

## Without Pipes

```{r}
#| eval: false
genetic_structure(arapat, stratum = "Species", mode = "Gst")
genetic_structure(arapat, stratum = "Species", mode = "Gst", nperm = 99)
pw <- genetic_structure(arapat, stratum = "Species", mode = "Gst", pairwise = TRUE)
genetic_structure(arapat, stratum = "Species", mode = "Gst", locus = c("LTRS", "WNT"))
freqs <- frequencies(arapat, stratum = "Species")
multilocus_assignment(arapat[1, ], freqs)
```
