# Population Graphs

Population Graphs are a graph-theoretic approach to understanding among-population genetic connectivity (Dyer & Nason 2004). Rather than assuming a specific model of gene flow (e.g., island or stepping-stone), population graphs estimate the conditional genetic covariance structure among populations, retaining only statistically significant edges.

```{r}
library(gstudio)
library(igraph)
data(arapat)
```

## Constructing a Population Graph

### From a Data Frame

The easiest way to create a population graph is with `population_graph()`, which handles the conversion from genotypes to multivariate data internally:

```{r}
graph <- population_graph(arapat, stratum = "Species")
graph
```

### From a Multivariate Matrix

For more control, convert genotypes to multivariate format first, then call `popgraph()` directly:

```{r}
mv <- to_mv(arapat)
groups <- factor(arapat$Species)
graph <- popgraph(mv, groups)
graph
```

The `alpha` parameter controls edge retention significance (default 0.05):

```{r}
graph_strict <- popgraph(mv, groups, alpha = 0.01)
graph_strict
```

### From an Adjacency Matrix

Use `as.popgraph()` to convert a pre-computed adjacency matrix:

```{r}
A <- matrix(c(0, 1, 0, 1,
              1, 0, 1, 0,
              0, 1, 0, 1,
              1, 0, 1, 0), nrow = 4)
rownames(A) <- colnames(A) <- LETTERS[1:4]
g <- as.popgraph(A)
g
```

## Bundled Population Graphs

The package includes two pre-computed population graphs:

```{r}
data(lopho)
lopho
```

```{r}
data(upiga)
upiga
```

## Decorating a Graph

The `decorate_graph()` function merges external data into graph vertex attributes. This is essential for spatial visualization:

```{r}
data(baja)
head(baja)
```

```{r}
lopho_dec <- decorate_graph(lopho, baja, stratum = "Population")
vertex_attr_names(lopho_dec)
```

Now the graph nodes carry coordinate and metadata attributes.

## Graph Metrics

### Node Properties

```{r}
# Vertex names
V(lopho)$name
```

```{r}
# Degree (number of edges)
degree(lopho)
```

```{r}
# Betweenness centrality
betweenness(lopho)
```

### Centroid Distance

The multivariate centroid distance between population centroids in discriminant space:

```{r}
mv <- to_mv(arapat)
groups <- factor(arapat$Species)
cd <- centroid_distance(mv, groups)
cd
```

### Centroid Variance

Multivariate variance around each population centroid:

```{r}
cv <- centroid_variance(mv, groups)
cv
```

### Shannon Entropy

Shannon entropy quantifies the uncertainty in a probability distribution:

```{r}
# Entropy of the degree distribution
deg <- degree(lopho)
p <- deg / sum(deg)
shannon_entropy(p)
```

### Edge Contortion

A measure of how edges deviate from geographic expectations (requires decorated graph with coordinates):

```{r}
#| eval: false
edge_contortion(lopho_dec)
```

## Congruence Testing

### Topology Congruence

Compare the topology of two population graphs:

```{r}
data(lopho)
data(upiga)
congruence_topology(lopho, upiga)
```

### Statistical Testing

Use `test_congruence()` to test whether two graphs are more similar in topology than expected by chance:

```{r}
#| eval: false
test_congruence(lopho, upiga, method = "combinatorial")
```

### Permutation

The `permute_popgraph()` function creates a null distribution of graph topologies by permuting the data:

```{r}
#| eval: false
null_graphs <- permute_popgraph(mv, groups, nperm = 99)
```

## Subgraph Operations

The `-` operator computes the difference between two popgraphs (removes edges in common). You can also extract subgraphs using igraph functions:

```{r}
smaller <- delete_vertices(lopho, "SenBas")
V(smaller)$name
```

## Export Formats

### To Adjacency Matrix

```{r}
M <- to_matrix(lopho)
M[1:5, 1:5]
```

### To Data Frame

```{r}
df <- to_df(lopho)
head(df)
```

### To JSON

```{r}
#| eval: false
json <- to_json(lopho)
cat(json)
```

### To Spatial Formats

```{r}
#| eval: false
# As sf object
sf_obj <- to_sf(lopho_dec)

# As KML
to_kml(lopho_dec, file = "lopho.kml")
```

### To pgraph Format

The `.pgraph` format is a simple text representation:

```{r}
#| eval: false
write_popgraph(lopho_dec, file = "lopho.pgraph")
```

### Reading a Population Graph

```{r}
path <- system.file("extdata", "lopho.pgraph", package = "gstudio")
g <- read_popgraph(path)
g
```

## Randomization

Generate a random graph with the same nodes and degree sequence as the original (the default, `mode = "degree"`, uses degree-preserving edge swaps), or with the same number of edges placed at random (`mode = "full"`):

```{r}
rg <- randomize_graph(lopho)
rg
```

## Genetic Gravity: Direction from an Undirected Graph

Each edge of a Population Graph is symmetric, but the two populations it joins are compared against different neighbourhoods. *Genetic gravity* reads direction from that difference. For population $i$, the neighbourhood weight $w_{j\mid i}$ is $j$'s share of a Gaussian kernel over $i$'s retained neighbours, with bandwidth $b_i = \gamma\, s_i/k_i$ (a multiple of $i$'s mean edge weight). The gravity asymmetry of an edge is $\Delta_{i\rightarrow j} = w_{i\mid j} - w_{j\mid i}$: positive when $i$ holds more weight in $j$'s neighbourhood than $j$ holds in $i$'s, marking $i$ as the genetic source. The source–sink score $S$ places every population on one scale, with $S_i - S_j \approx \Delta_{i\rightarrow j}$ over all edges. Throughout, **higher means more of a source**.

### Choosing the bandwidth

Two bandwidths answer different questions:

- `gamma = 0.5`, the *degree-neutral* bandwidth (the default in the functions below), for **inferring direction**. At this bandwidth, source–sink scores carry essentially no degree signature under symmetric migration.
- `gamma = 1`, the *local mean*, for **describing connectivity**. It keeps the contribution of degree, so hubs read as sources and leaves as sinks.

`gamma = Inf` gives the flat kernel, where every quantity is set by degree alone. Whatever the bandwidth, report the *degree share*: the share of the variance in $\Delta$ explained by the degree-only term $\Delta^0_{i\rightarrow j} = 1/k_j - 1/k_i$.

```{r}
pops  <- population_graph(arapat, stratum = "Population")
coords <- strata_coordinates(arapat)

scores <- source_sink_scores(pops, gamma = 0.5)
head(scores[order(-scores$S), c("node", "S", "gravity", "degree")])
c(degree_share = attr(scores, "degree_share"), cor_S_degree = attr(scores, "cor_S_degree"))

# the same graph at the local mean: degree carries much more of the index
attr(source_sink_scores(pops, gamma = 1), "degree_share")
```

Edge-level quantities, in the same sign convention, come from `gravity_edges()`. The neighbourhood-weight matrix comes from `neighbourhood_weights()`:

```{r}
head(gravity_edges(pops))
```

### Mapping sources and sinks

`plot_gravity_field()` interpolates $S$ over the node coordinates (red = source, blue = sink), draws the edges, and places an arrow at each edge midpoint, pointing from source to sink with length proportional to $|\Delta|$. A named list of graphs, or one graph at several bandwidths, is drawn on common scales. `mask_dist` sets how far from a node the surface extends (default 0.75 times the median nearest-neighbour distance); clustered sampling locations usually need a larger value:

```{r}
#| fig-width: 9
#| fig-height: 5
plot_gravity_field(list(`local mean (gamma = 1)` = pops, `degree-neutral (gamma = 1/2)` = pops),
                   coords = coords, gamma = c(1, 0.5), mask_dist = 0.6)
```

The quantities behind the map are a `gravity_field` object. Compute it once, then print it, convert it with `as.data.frame()` (nodes or edges) for your own plots, or `plot()` it. A list of fields, or of fields and graphs mixed, can be passed to `plot_gravity_field()`; each field keeps its own coordinates and bandwidth:

```{r}
field <- gravity_field(pops, coords = coords, gamma = 0.5)
field
head(as.data.frame(field, what = "edges"))
```

### Testing for a directional gradient

`source_sink_test()` asks whether $S$ changes monotonically along a hypothesized covariate, such as position along a river or, here, latitude along the peninsula. The statistic is Spearman's $r(S, x)$; $r < 0$ places the sources at low $x$. The default null is Moran spectral randomization on the graph's own topology (`msr_null()`, native to gstudio). It reproduces the node-level smoothness that $S$ inherits from drift; a plain permutation of $S$ does not, and is anti-conservative.

```{r}
lat <- setNames(coords$Latitude, coords$Stratum)
set.seed(1)
source_sink_test(pops, x = lat, nperm = 499)
```

### Directional isolation by graph distance

`pgd()` splits each edge's conditional genetic distance into two directional parts in proportion to the neighbourhood weights. Because it is a distance, it is shorter in the direction of gene flow. `directional_ibgd()` asks whether isolation by graph distance is described better by these directed distances, with separate forward and reverse slopes along the hypothesized axis, than by the symmetric cGD. $\Delta R^2 > 0$ favours the directional description; the slope ratio $b_{fwd}/b_{rev} < 1$ means distance accumulates more slowly downstream.

```{r}
directional_ibgd(pops, x = lat)
```

The two tests answer different questions. Under strong drift the model comparison can prefer a directional fit by chance, so read it alongside the source–sink test. Edge-level permutation tests of $\Delta$ (`asymmetry_significance()` and its modes) are not recommended as tests of direction: under symmetric migration they are anti-conservative, or never reject.

### Data quality: information loss

As loci fix, graphs are estimated from fewer informative loci, which inflates $|\Delta|$ whatever the direction of gene flow. `locus_subsample_null()` rebuilds the graph from random subsets of polymorphic loci, to show how much $|\Delta|$ information loss alone can produce at a given level of polymorphism:

```{r}
#| eval: false
locus_subsample_null(arapat, k = c(2, 3, 4, 6), nsub = 3)
```

### Simulating directional gene flow

A directional stepping stone, with rates to the next population downstream (`m_fwd`) and upstream (`m_rev`), plugs into `simulate_pop()`:

```{r}
#| eval: false
f <- data.frame(Population = rep(sprintf("Pop%02d", 1:10), each = 2), Locus = "L1",
                Allele = rep(c("01", "02"), 10), Frequency = 0.5)
pop <- make_populations(f, N = 50)
mm  <- migration_matrix(sprintf("Pop%02d", 1:10), model = "stepping_stone_1d_asym",
                        m_fwd = 0.04, m_rev = 0.01)
out <- simulate_pop(pop, ngen = 200, migration = migration_event(mm), verbose = FALSE)
```
