# Simulation & Parentage

This chapter covers the forward-time population simulation framework and parentage analysis tools in `gstudio`.

```{r}
#| message: false
library(gstudio)
library(dplyr)
```

## Building Populations from Frequencies

### Allele Frequency Specification

The simulation tools start with specifying allele frequencies per population and locus. The `make_populations()` function requires a data.frame with `Population`, `Locus`, `Allele`, and `Frequency` columns:

```{r}
freqs <- expand.grid(Allele = c("01", "02"),
                     Locus = c("Loc1", "Loc2", "Loc3"),
                     Population = c("A", "B", "C"),
                     stringsAsFactors = FALSE) |>
  mutate(Frequency = c(0.2, 0.8, 0.3, 0.7, 0.4, 0.6,    # A
                       0.7, 0.3, 0.5, 0.5, 0.6, 0.4,    # B
                       0.4, 0.6, 0.3, 0.7, 0.2, 0.8),   # C
         Population = factor(Population, ordered = TRUE)) |>
  select(Population, Locus, Allele, Frequency)
```

### Creating Populations

```{r}
pop <- freqs |>
  make_populations(N = 30)

pop |>
  head()

pop |>
  count(Population)
```

You can specify an inbreeding level with `F`:

```{r}
pop_inbred <- freqs |>
  make_populations(N = 30, F = 0.3)
```

## Migration Models

The `migration_matrix()` function creates a K x K migration matrix where rows sum to 1. Element `[i,j]` is the proportion of population *i* that originates from population *j*.

### Island Model

Equal migration among all populations:

```{r}
migration_matrix(3, model = "island", m = 0.05)
```

A count names the populations `Pop1`, `Pop2`, ….  For simulations, pass the population names instead (as below), so the matrix matches the `Population` column that `simulate_pop()` migrates between.

### 1D Stepping Stone

Migration only between adjacent populations:

```{r}
c("A", "B", "C", "D") |>
  migration_matrix(model = "stepping_stone_1d", m = 0.1)
```

### 2D Stepping Stone

Grid-based adjacency:

```{r}
migration_matrix(4, model = "stepping_stone_2d",
                 m = 0.1, nr = 2, nc = 2)
```

### Distance-Based

Migration rate decays with distance:

```{r}
coords <- matrix(c(0, 0, 1, 0, 0.5, 1), ncol = 2, byrow = TRUE)
migration_matrix(3, model = "distance", m = 0.1, coords = coords)
```

## Migration Events

A `migration_event` pairs a migration matrix with a generation interval, allowing temporal regime changes:

```{r}
# Low migration for generations 1-50
evt1 <- c("A", "B", "C") |>
  migration_matrix(model = "island", m = 0.05) |>
  migration_event(start = 1, end = 50)

# High migration from generation 51 onward
evt2 <- c("A", "B", "C") |>
  migration_matrix(model = "island", m = 0.20) |>
  migration_event(start = 51)
```

## Mutation Models

The `mutation_model()` function defines how alleles mutate during simulation:

```{r}
# Infinite Alleles Model
mm_iam <- mutation_model(rate = 0.001, model = "iam")

# K-Allele Model (10 possible states)
mm_kam <- mutation_model(rate = 0.001, model = "kam", k = 10)

# Stepwise Mutation Model
mm_smm <- mutation_model(rate = 0.001, model = "smm")
```

## Forward-Time Simulation

The `simulate_pop()` function runs a multi-generation forward-time simulation. Each generation cycle: census, migrate, mate, mutate.

### Basic Simulation

```{r}
result <- pop |>
  simulate_pop(ngen = 20, verbose = FALSE)

result |>
  head()
```

### With Migration

```{r}
evt <- pop |>
  pull(Population) |>
  unique() |>
  migration_matrix(model = "island", m = 0.05) |>
  migration_event(start = 1)

result <- pop |>
  simulate_pop(ngen = 20, migration = evt, verbose = FALSE)
```

### With Mutation

```{r}
mm <- mutation_model(rate = 0.01, model = "iam")

result <- pop |>
  simulate_pop(ngen = 20, mutation = mm, verbose = FALSE)
```

### With Mixed Mating (Selfing)

```{r}
result <- pop |>
  simulate_pop(ngen = 20, selfing_rate = 0.3, verbose = FALSE)
```

### Census Snapshots

Save population snapshots at regular intervals:

```{r}
#| eval: false
result <- pop |>
  simulate_pop(ngen = 100, census_interval = 25, census_dir = tempdir(), verbose = FALSE)

# Snapshots saved as gen_0025.rds, gen_0050.rds, etc.
```

### Complete Worked Example

Simulate drift and migration in a 3-population system, then analyze the outcome:

```{r}
# Create starting populations
pop <- freqs |>
  make_populations(N = 50)

# Set up island model migration among the simulated populations
evt <- pop |>
  pull(Population) |>
  unique() |>
  migration_matrix(model = "island", m = 0.02) |>
  migration_event(start = 1)

# Run simulation
set.seed(42)
final <- pop |>
  simulate_pop(ngen = 50, migration = evt, verbose = FALSE)

# Analyze structure in the final generation
final |>
  genetic_structure(stratum = "Population", mode = "Gst")
```

```{r}
# Compare diversity before and after
list(Start = pop, Final = final) |>
  lapply(genetic_diversity, stratum = "Population", mode = "He") |>
  bind_rows(.id = "Generation") |>
  group_by(Generation, Stratum) |>
  summarize(mean_He = mean(He), .groups = "drop")
```

## Parentage Analysis

### Mating and Offspring

The `mate()` function creates offspring from two parents:

```{r}
f <- data.frame(
  Locus = rep(paste0("Loc", 1:4), each = 3),
  Allele = rep(LETTERS[1:3], times = 4),
  Frequency = rep(c(1/3, 1/3, 1/3), times = 4)
)
adults <- f |>
  make_population(N = 20)

offs <- adults |>
  slice(1) |>
  mate(slice(adults, 2), N = 5)

offs
```

### Maternal Subtraction

The `minus_mom()` function removes the maternal contribution from offspring genotypes. It takes a single data.frame where mothers have `OffID = 0` and offspring have `OffID != 0`:

```{r}
# Combine mother (OffID = 0) and offspring into one data.frame
family <- bind_rows(adults |>
                      slice(1) |>
                      mutate(OffID = 0),
                    offs |>
                      mutate(OffID = row_number()))

family |>
  minus_mom()
```

### Fractional Paternity

The `paternity()` function estimates the probability of each potential father siring each offspring:

```{r}
offs <- offs |>
  mutate(OffID = row_number(),
         MomID = adults$ID[1])

pat <- offs |>
  paternity(slice(adults, 1), adults)

pat |>
  head()
```

The `Fij` column gives the fractional paternity probability. Use `strict = TRUE` to retain only unambiguous assignments:

```{r}
offs |>
  paternity(slice(adults, 1), adults, strict = TRUE)
```

### Single-Parent Exclusion

The `parent_finder()` function identifies compatible parents for each offspring using exclusion. It requires a data.frame with `ID` and `OffID` columns, where adults have `OffID = 0`:

```{r}
# Prepare a combined data.frame: adults (OffID = 0) then offspring
locus_cols <- adults |>
  column_class("locus")

combined <- bind_rows(adults |>
                        select(ID, all_of(locus_cols)) |>
                        mutate(OffID = 0),
                      offs |>
                        select(ID, all_of(locus_cols)) |>
                        mutate(OffID = row_number()))

combined |>
  parent_finder() |>
  head()
```

### Identifying Bad Parents

The `bad_parents()` function flags parent-offspring pairs that are genetically incompatible:

```{r}
#| eval: false
offs |>
  bad_parents(adults)
```

## Bundled Parentage Datasets

The `cornus` and `cornus_florida` datasets contain *Cornus florida* genotype data for parentage analysis:

```{r}
data(cornus)

cornus |>
  head()
```

```{r}
data(cornus_florida)

cornus_florida |>
  head()
```

## Without Pipes

```{r}
#| eval: false
pop    <- make_populations(freqs, N = 50)
m      <- migration_matrix(unique(pop$Population), model = "island", m = 0.02)
evt    <- migration_event(m, start = 1)
final  <- simulate_pop(pop, ngen = 50, migration = evt, verbose = FALSE)
genetic_structure(final, stratum = "Population", mode = "Gst")

adults <- make_population(f, N = 20)
offs   <- mate(adults[1, ], adults[2, ], N = 5)
pat    <- paternity(offs, adults[1, ], adults)
```
