# Visualization

The `gstudio` package returns analysis results as standard `data.frame` objects, which plug directly into ggplot2. This chapter shows the recommended plotting patterns for allele frequency data, population maps, and population graphs.

```{r}
library(gstudio)
library(ggplot2)
library(ggraph)
data(arapat)
```

## Allele Frequency Plots

`frequencies()` returns a tidy data frame with columns `Locus`, `Allele`, and `Frequency` (plus `Stratum` when partitioned). Pass it directly to `geom_bar()`.

### Single locus

```{r}
freqs <- frequencies(arapat, loci = "LTRS")
ggplot(freqs, aes(x = Allele, y = Frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Allele Frequencies at LTRS") +
  theme_minimal()
```

### Partitioned by stratum

When `stratum` is supplied, `frequencies()` zero-fills missing allele/stratum combinations so dodged and faceted plots are correctly aligned. Filter with `Frequency > 0` beforehand if you prefer to show only observed alleles.

```{r}
freqs <- frequencies(arapat, loci = "LTRS", stratum = "Species")
ggplot(freqs, aes(x = Allele, y = Frequency, fill = Stratum)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "LTRS Allele Frequencies by Species") +
  theme_minimal()
```

### Faceted across strata

```{r}
ggplot(freqs, aes(x = Allele, y = Frequency, fill = Allele)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Stratum) +
  labs(title = "LTRS Frequencies by Species") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Spatial Sampling Maps

`strata_coordinates()` aggregates individual records to population centroids, returning a data frame with `Stratum`, `Longitude`, and `Latitude` columns ready for `geom_point()`.

```{r}
coords <- strata_coordinates(arapat)
ggplot(coords, aes(x = Longitude, y = Latitude)) +
  geom_point(size = 3, shape = 21, fill = "steelblue") +
  labs(title = "Araptus attenuatus Sampling Locations") +
  coord_fixed() +
  theme_minimal()
```

Individual-level data works just as well when you want within-population spread:

```{r}
ggplot(arapat, aes(x = Longitude, y = Latitude, fill = Species)) +
  geom_point(size = 2, shape = 21, alpha = 0.7) +
  coord_fixed() +
  theme_minimal()
```

## Population Graph Visualization

Population graphs are best visualized with the [ggraph](https://ggraph.data-imaginist.com) package, which extends ggplot2 for graph objects and works natively with `popgraph`/`igraph` objects. Install it once with `install.packages("ggraph")`.

### Force-Directed Layout

Build a population graph from the `arapat` dataset and plot it with a Fruchterman-Reingold layout. Node and edge aesthetics map directly to vertex and edge attributes.

```{r}
graph <- population_graph(arapat, stratum = "Species")
igraph::V(graph)$degree <- igraph::degree(graph)

ggraph(graph, layout = "fr") +
  geom_edge_link(color = "grey60", linewidth = 0.5) +
  geom_node_point(aes(fill = degree), size = 5, shape = 21) +
  geom_node_label(aes(label = name), repel = TRUE, size = 2.5) +
  scale_fill_viridis_c() +
  theme_graph() +
  labs(title = "Araptus attenuatus Population Graph")
```

### Geographic (IBD) Layout

For isolation-by-distance analyses, use `layout = "manual"` with longitude and latitude pulled from vertex attributes set by `decorate_graph()`. Add `coord_fixed()` to preserve geographic proportions.

Use `scale_edge_width(range = c(min, max))` to control the rendered line width in mm. Dividing weights inside `aes()` has no effect because ggraph's default scale auto-normalises to the data range regardless — the scale absorbs any constant factor. For non-linear transforms (log, normalize), add a pre-computed edge attribute with `igraph::E(graph)$attr <- ...` before plotting.

```{r}
data(lopho)
data(baja)
lopho_dec <- decorate_graph(lopho, baja, stratum = "Population")

ggraph(lopho_dec, layout = "manual",
       x = igraph::V(lopho_dec)$Longitude,
       y = igraph::V(lopho_dec)$Latitude) +
  geom_edge_link(aes(width = weight), color = "grey60", show.legend = FALSE) +
  scale_edge_width(range = c(0.3, 2.5), guide = "none") +
  geom_node_point(aes(size = size, color = Region), shape = 21) +
  geom_node_label(aes(label = name), repel = TRUE, size = 2.5) +
  coord_fixed() +
  labs(title = "Lophocereus Population Graph",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

### Directed (Asymmetric) Graphs

`asymmetric_popgraph()` produces a directed graph where each undirected edge becomes two directed edges with weights reflecting the net introgression pressure. `geom_edge_fan()` automatically curves parallel edges so both directions are visible.

```{r}
lopho_asym <- asymmetric_popgraph(lopho_dec)

ggraph(lopho_asym, layout = "manual",
       x = igraph::V(lopho_asym)$Longitude,
       y = igraph::V(lopho_asym)$Latitude) +
  geom_edge_fan(aes(color = weight, width = weight),
                arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
                end_cap = circle(3, "mm"),
                show.legend = FALSE) +
  scale_edge_width(range = c(0.3, 2), guide = "none") +
  scale_edge_color_viridis(option = "magma", alpha = 0.4) +
  geom_node_point(size = 3, shape = 21, fill = "white", stroke = 0.8) +
  coord_fixed() +
  labs(title = "Asymmetric Gene Flow — Lophocereus",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

## Interactive Leaflet Maps

`addAlleleFrequencies()` adds allele-frequency pie charts to a leaflet map and follows the standard leaflet pipe pattern. Pre-compute frequencies with `frequencies()`, build the base map from `strata_coordinates()`, then pipe into `addAlleleFrequencies()`.

```{r}
library(leaflet)
library(leaflet.minicharts)

freqs <- frequencies(arapat, loci = "LTRS", stratum = "Species")

strata_coordinates(arapat, stratum = "Species") |>
  leaflet() |>
  addTiles() |>
  addAlleleFrequencies(freqs)
```

Pass `locus =` explicitly when `freqs` contains more than one locus. Width and height of the pie charts (in pixels) are controlled with `width =` and `height =`.
