# Genetic Distance & Relatedness

This chapter covers inter-individual and inter-population distance estimation, pairwise relatedness, and spatial autocorrelation analysis.

```{r}
#| message: false
library(gstudio)
library(dplyr)
data(arapat)
```

## Genetic Distance

The `genetic_distance()` function computes distance matrices from genetic data. Some modes operate among individuals, others among populations (strata).

### Available Distance Metrics

| Mode | Level | Description |
|------|-------|-------------|
| `"AMOVA"` | Individual | Inter-individual AMOVA distance |
| `"Bray"` | Individual | Proportion of shared alleles |
| `"Euclidean"` | Population | Euclidean frequency distance |
| `"cGD"` | Population | Conditional genetic distance |
| `"pGD"` | Population | Partitioned (directed) conditional genetic distance; asymmetric |
| `"Nei"` | Population | Nei's genetic distance (1978) |
| `"Dps"` | Population | Shared allele distance (1 - Ps) |
| `"Jaccard"` | Population | Jaccard set dissimilarity |

### Individual-Level Distances

AMOVA distance computes inter-individual genetic distances:

```{r}
# The first 20 individuals, for a manageable example
D_amova <- arapat |>
  slice(1:20) |>
  genetic_distance(mode = "AMOVA")

dim(D_amova)
D_amova[1:5, 1:5]
```

### Population-Level Distances

Population-level distances require a stratum column:

```{r}
D_nei <- arapat |>
  genetic_distance(stratum = "Species", mode = "Nei")
D_nei
```

```{r}
D_cgd <- arapat |>
  genetic_distance(stratum = "Species", mode = "cGD")
D_cgd
```

Partitioned conditional genetic distance (pGD) splits each Population Graph edge's cGD into two directed shares using the genetic-gravity neighbourhood weights, so the matrix is asymmetric: `[i, j]` is the distance from `i` to `j`, shorter in the direction of gene flow. The bandwidth multiplier is set with `gamma` (default 0.5); see `vignette("genetic_gravity")`.

```{r}
D_pgd <- arapat |>
  genetic_distance(stratum = "Species", mode = "pGD")
D_pgd
```

```{r}
D_euc <- arapat |>
  genetic_distance(stratum = "Species", mode = "Euclidean")
D_euc
```

## Genetic Relatedness

The `genetic_relatedness()` function computes pairwise relatedness among individuals.

### Available Relatedness Estimators

| Mode | Description |
|------|-------------|
| `"Nason"` | Nason's kinship coefficient (Fij) |
| `"LynchRitland"` | Lynch & Ritland estimator |
| `"Ritland"` | Ritland estimator |

### Basic Usage

The Nason estimator divides by the polymorphic index `Pe()`, so monomorphic loci (where `Pe = 0`) will produce `NaN`. In the `arapat` dataset, `LTRS` and `ATPS` can be monomorphic in small subsets. To avoid this, restrict the data to polymorphic loci:

```{r}
# Use loci that are polymorphic in this subset
R <- arapat |>
  slice(1:10) |>
  select(Species, WNT, EN, EF, ZMP, AML, MP20) |>
  genetic_relatedness(mode = "Nason")

dim(R)
round(R[1:5, 1:5], 4)
```

### Single-Locus Relatedness

Relatedness at a single locus uses the same function, with `loci =`:

```{r}
r_nason <- arapat |>
  slice(1:10) |>
  genetic_relatedness(loci = "EN", mode = "Nason")

round(r_nason[1:5, 1:5], 4)
```

## Spatial Autocorrelation

The `genetic_autocorrelation()` function performs the spatial autocorrelation analysis of Smouse & Peakall (1999). It requires two square distance matrices --- one physical, one genetic --- and a vector of distance bin boundaries:

```{r}
#| eval: false
# Physical distance matrix
P <- arapat |>
  strata_distance(longitude = "Longitude", latitude = "Latitude")

# Genetic distance matrix
G <- arapat |>
  genetic_distance(mode = "AMOVA")

# Define distance bins
bins <- seq(0, max(P), length.out = 6)

# Run autocorrelation
result <- genetic_autocorrelation(P, G, bins = bins, perms = 99)
result
```

The output includes the spatial autocorrelation coefficient (R) for each distance class, the number of pairwise comparisons (N), and a p-value from the permutation test (P).

## Strata Distances

The `strata_coordinates()` function extracts per-stratum centroid coordinates, and `strata_distance()` computes physical (geographic) distances between them:

```{r}
coords <- arapat |>
  strata_coordinates(stratum = "Species", longitude = "Longitude", latitude = "Latitude")
coords
```

```{r}
coords |>
  strata_distance()
```

## Without Pipes

```{r}
#| eval: false
D_amova <- genetic_distance(arapat[1:20, ], mode = "AMOVA")
D_nei   <- genetic_distance(arapat, stratum = "Species", mode = "Nei")
D_pgd   <- genetic_distance(arapat, stratum = "Species", mode = "pGD", gamma = 0.5)
R       <- genetic_relatedness(arapat[1:10, c("Species", "WNT", "EN", "EF", "ZMP", "AML", "MP20")],
                               mode = "Nason")
coords  <- strata_coordinates(arapat, stratum = "Species")
strata_distance(coords)
```
