# Getting Started

This chapter covers the installation of `gstudio`, loading bundled datasets, and a quick tour of the package's core utilities.

## Installation

The `gstudio` package is available from GitHub and can be installed using the `devtools` package:

```{r}
#| eval: false
install.packages("devtools")
devtools::install_github("dyerlab/gstudio")
```

Once installed, load the package:

```{r}
library(gstudio)
```

## Bundled Datasets

The package ships with several datasets for examples and testing:

| Dataset | Description |
|---------|-------------|
| `arapat` | *Araptus attenuatus* multilocus microsatellite data across Baja California |
| `aflp_arapat` | *Araptus attenuatus* AFLP marker data |
| `cornus` | *Cornus florida* adult genotypes |
| `cornus_florida` | *Cornus florida* mother-offspring pair data |
| `lopho` | *Lophocereus schottii* population graph (Dyer & Nason 2004) |
| `upiga` | *Upiga virescens* population graph |
| `baja` | Baja California sampling location metadata |
| `alt` | Sonoran desert altitude raster data |

Load any dataset with the `data()` function:

```{r}
data(arapat)
```

## Exploring a Dataset

The `arapat` dataset is a `data.frame` with both metadata columns and `locus` columns:

```{r}
names(arapat)
dim(arapat)
```

### Identifying Locus Columns

Use `column_class()` to find all columns of a particular class:

```{r}
column_class(arapat, "locus")
```

This returns the names of all columns that contain `locus` objects --- a function used extensively throughout the package.

### Splitting by Stratum

The `partition()` function splits a data.frame by a grouping column, returning a named list:

```{r}
pops <- partition(arapat, stratum = "Species")
names(pops)
sapply(pops, nrow)
```

### Ploidy

Check the ploidy level of a locus column:

```{r}
ploidy(arapat$LTRS)
```

## A Quick Analysis

With just a few lines, you can estimate genetic diversity across populations:

```{r}
genetic_diversity(arapat, stratum = "Species", mode = "He")
```

Or compute population structure:

```{r}
genetic_structure(arapat, stratum = "Species", mode = "Gst")
```

The following chapters cover each of these operations in detail.
