# Getting Started

This chapter covers the installation of `gstudio`, loading bundled datasets, and a quick tour of the package's core utilities.

## Installation

The `gstudio` package is available from GitHub and can be installed using the `devtools` package:

```{r}
#| eval: false
install.packages("devtools")
devtools::install_github("dyerlab/gstudio")
```

Once installed, load the package, along with dplyr for the pipelines used throughout this book:

```{r}
#| message: false
library(gstudio)
library(dplyr)
```

## Code Style in This Book

Every `gstudio` function takes its data as the first argument, so analyses are written as pipelines with the native pipe `|>`, one function per line:

```{r}
#| eval: false
arapat |>
  genetic_diversity(stratum = "Species", mode = "He")
```

The older style, with the data and every option passed as arguments, works exactly the same and needs no other packages:

```{r}
#| eval: false
genetic_diversity(arapat, stratum = "Species", mode = "He")
```

Grouped data frames work too: when no `stratum` is given, the grouping variable is used, so these two calls are the same (an explicit `stratum =` always takes precedence over the grouping):

```{r}
#| eval: false
arapat |>
  group_by(Species) |>
  genetic_diversity(mode = "He")

arapat |>
  genetic_diversity(stratum = "Species", mode = "He")
```

Tests (`source_sink_test()`, `ibgd()`, `directional_test()`, the graph-level `asymmetry_significance()` and others) return standard `htest` objects, so they print like `t.test()` and can be turned into a one-row data frame with `broom::tidy()`.

Each chapter uses pipelines in its running examples and ends with a short, non-evaluated *Without pipes* block giving the argument-style equivalents. The help pages (`?genetic_diversity` and so on) use the argument style throughout.

## Bundled Datasets

The package ships with several datasets for examples and testing:

| Dataset | Description |
|---------|-------------|
| `arapat` | *Araptus attenuatus* multilocus microsatellite data across Baja California |
| `aflp_arapat` | *Araptus attenuatus* AFLP marker data |
| `cornus` | *Cornus florida* adult genotypes |
| `cornus_florida` | *Cornus florida* mother-offspring pair data |
| `lopho` | *Lophocereus schottii* population graph (Dyer & Nason 2004) |
| `upiga` | *Upiga virescens* population graph |
| `baja` | Baja California sampling location metadata |
| `gravity` | Simulated genotypes under directional gene flow (Genetic Gravity) |

Load any dataset with the `data()` function:

```{r}
data(arapat)
```

## Exploring a Dataset

The `arapat` dataset is a `data.frame` with both metadata columns and `locus` columns:

```{r}
arapat |>
  names()

arapat |>
  dim()
```

### Identifying Locus Columns

Use `column_class()` to find all columns of a particular class:

```{r}
arapat |>
  column_class("locus")
```

This returns the names of all columns that contain `locus` objects --- a function used extensively throughout the package.

### Splitting by Stratum

The `partition()` function splits a data.frame by a grouping column, returning a named list:

```{r}
pops <- arapat |>
  partition(stratum = "Species")

pops |>
  sapply(nrow)
```

With dplyr, the same counts come from grouping:

```{r}
arapat |>
  count(Species)
```

### Ploidy

Check the ploidy level of a locus column:

```{r}
arapat |>
  pull(LTRS) |>
  ploidy()
```

## A Quick Analysis

With just a few lines, you can estimate genetic diversity across populations:

```{r}
arapat |>
  genetic_diversity(stratum = "Species", mode = "He")
```

Or compute population structure:

```{r}
arapat |>
  genetic_structure(stratum = "Species", mode = "Gst")
```

The following chapters cover each of these operations in detail.

## Without Pipes

```{r}
#| eval: false
names(arapat)
column_class(arapat, "locus")
pops <- partition(arapat, stratum = "Species")
sapply(pops, nrow)
ploidy(arapat$LTRS)
genetic_diversity(arapat, stratum = "Species", mode = "He")
genetic_structure(arapat, stratum = "Species", mode = "Gst")
```
