# Population Graphs

Population Graphs are a graph-theoretic approach to understanding among-population genetic connectivity (Dyer & Nason 2004). Rather than assuming a specific model of gene flow (e.g., island or stepping-stone), population graphs estimate the conditional genetic covariance structure among populations, retaining only statistically significant edges.

```{r}
#| message: false
library(gstudio)
library(dplyr)
library(igraph)
data(arapat)
```

## Constructing a Population Graph

### From a Data Frame

The easiest way to create a population graph is with `population_graph()`, which handles the conversion from genotypes to multivariate data internally:

```{r}
graph <- arapat |>
  population_graph(stratum = "Species")
graph
```

### From a Multivariate Matrix

For more control, convert genotypes to multivariate format first, then call `popgraph()` directly:

```{r}
mv <- arapat |>
  to_mv()

groups <- arapat |>
  pull(Species) |>
  factor()

graph <- mv |>
  popgraph(groups)

graph
```

The `alpha` parameter controls edge retention significance (default 0.05):

```{r}
graph_strict <- mv |>
  popgraph(groups, alpha = 0.01)
graph_strict
```

### From an Adjacency Matrix

Use `as.popgraph()` to convert a pre-computed adjacency matrix:

```{r}
A <- matrix(c(0, 1, 0, 1,
              1, 0, 1, 0,
              0, 1, 0, 1,
              1, 0, 1, 0), nrow = 4)
rownames(A) <- colnames(A) <- LETTERS[1:4]
g <- A |>
  as.popgraph()
g
```

## Bundled Population Graphs

The package includes two pre-computed population graphs:

```{r}
data(lopho)
lopho
```

```{r}
data(upiga)
upiga
```

## Decorating a Graph

The `decorate_graph()` function merges external data into graph vertex attributes. This is essential for spatial visualization:

```{r}
data(baja)
baja |>
  head()
```

```{r}
lopho_dec <- lopho |>
  decorate_graph(baja, stratum = "Population")
lopho_dec |>
  vertex_attr_names()
```

Now the graph nodes carry coordinate and metadata attributes.

## Graph Metrics

### Node Properties

```{r}
# Vertex names
V(lopho)$name
```

```{r}
# Degree (number of edges)
lopho |>
  degree()
```

```{r}
# Betweenness centrality
lopho |>
  betweenness()
```

### Centroid Distance

The multivariate centroid distance between population centroids in discriminant space:

```{r}
cd <- mv |>
  centroid_distance(groups)

cd
```

### Centroid Variance

Multivariate variance around each population centroid:

```{r}
cv <- mv |>
  centroid_variance(groups)
cv
```

### Shannon Entropy

Shannon entropy quantifies the uncertainty in a probability distribution:

```{r}
# Entropy of the degree distribution
deg <- lopho |>
  degree()

(deg / sum(deg)) |>
  shannon_entropy()
```

### Edge Contortion

A measure of how edges deviate from geographic expectations (requires decorated graph with coordinates):

```{r}
#| eval: false
lopho_dec |>
  edge_contortion()
```

## Congruence Testing

### Topology Congruence

Compare the topology of two population graphs:

```{r}
data(lopho)
data(upiga)
lopho |>
  congruence_topology(upiga)
```

### Statistical Testing

Use `test_congruence()` to test whether two graphs are more similar in topology than expected by chance:

```{r}
#| eval: false
lopho |>
  test_congruence(upiga, method = "combinatorial")
```

### Permutation

The `permute_popgraph()` function creates a null distribution of graph topologies by permuting the data:

```{r}
#| eval: false
null_graphs <- mv |>
  permute_popgraph(groups, nperm = 99)
```

## Subgraph Operations

The `-` operator computes the difference between two popgraphs (removes edges in common). You can also extract subgraphs using igraph functions:

```{r}
smaller <- lopho |>
  delete_vertices("SenBas")
V(smaller)$name
```

## Export Formats

### To Adjacency Matrix

```{r}
M <- lopho |>
  to_matrix()
M[1:5, 1:5]
```

### To Data Frame

```{r}
lopho |>
  to_df() |>
  head()
```

### To JSON

```{r}
#| eval: false
lopho |>
  to_json() |>
  cat()
```

### To Spatial Formats

```{r}
#| eval: false
# As sf object
sf_obj <- lopho_dec |>
  to_sf()

# As KML
lopho_dec |>
  to_kml(file = "lopho.kml")
```

### To pgraph Format

The `.pgraph` format is a simple text representation:

```{r}
#| eval: false
lopho_dec |>
  write_popgraph(file = "lopho.pgraph")
```

### Reading a Population Graph

```{r}
g <- system.file("extdata", "lopho.pgraph", package = "gstudio") |>
  read_popgraph()

g
```

## Plotting

`plot()` draws a Population Graph with ggplot2 and returns a `ggplot` you can extend with `+`. Nodes are placed from `Longitude`/`Latitude` vertex attributes when present (for example after `decorate_graph()`), and otherwise with a Kamada-Kawai layout; pass `layout =` to choose another layout name, a layout function, or a coordinate table. The same `layout`, `node_size`, `node_labels`, `node_fill`, `edges` and `edge_width` options apply to `plot()` on a gravity field and to `animate_popgraphs()`. Directed graphs from `asymmetric_popgraph()` are drawn with arrows. For fully custom figures, use ggraph (see the Visualization article).

```{r}
lopho |>
  plot()
```

## Randomization

Generate a random graph with the same nodes and degree sequence as the original (the default, `mode = "degree"`, uses degree-preserving edge swaps), or with the same number of edges placed at random (`mode = "full"`):

```{r}
rg <- lopho |>
  randomize_graph()
rg
```

## Genetic Gravity: Direction from an Undirected Graph

Each edge of a Population Graph is symmetric, but the two populations it joins are compared against different neighbourhoods. *Genetic gravity* reads the direction of gene flow from that difference: edge asymmetries $\Delta$, a source–sink score $S$ for every population, the partitioned distance pGD, tests for a directional gradient, and maps of the gravity field. It has its own chapter, @sec-genetic-gravity, which walks through the analysis from raw genotypes.

```{r}
lopho |>
  gravity_field() |>
  plot()
```

## Without Pipes

```{r}
#| eval: false
graph     <- population_graph(arapat, stratum = "Species")
graph     <- popgraph(to_mv(arapat), factor(arapat$Species), alpha = 0.05)
lopho_dec <- decorate_graph(lopho, baja, stratum = "Population")
cd        <- centroid_distance(to_mv(arapat), factor(arapat$Species))
congruence_topology(lopho, upiga)
M         <- to_matrix(lopho)
g         <- read_popgraph(system.file("extdata", "lopho.pgraph", package = "gstudio"))
plot(lopho, layout = "kk")
plot(gravity_field(lopho))
```
